---
output:
  rmarkdown::github_document:
    html_preview: no
---

<!--- README.md is generated from README.Rmd. Please edit that file -->

# raptr: Representative and Adequate Prioritization Toolkit in R

[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![Travis Build Status](http://img.shields.io/travis/jeffreyhanson/raptr/master.svg?label=Mac%20OSX%20%26%20Linux)](https://travis-ci.org/jeffreyhanson/raptr)
[![AppVeyor Build Status](http://img.shields.io/appveyor/ci/jeffreyhanson/raptr/master.svg?label=Windows)](https://ci.appveyor.com/project/jeffreyhanson/raptr)
[![Coverage Status](http://codecov.io/github/jeffreyhanson/raptr/coverage.svg?branch=master)](https://codecov.io/github/jeffreyhanson/raptr?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/raptr)](https://CRAN.R-project.org/package=raptr)

Biodiversity is in crisis. The overarching aim of conservation is to preserve biodiversity patterns and processes. To this end, protected areas are established to buffer species and preserve biodiversity processes. But resources are limited and so protected areas must be cost-effective. This package contains tools to generate plans for protected areas (prioritizations), using spatially explicit targets for biodiversity patterns and processes. To obtain solutions in a feasible amount  of time, this package uses the commercial 'Gurobi' software package (obtained from http://www.gurobi.com/). Additionally, the 'rgurobi' package can also be installed to provide extra functionality (obtained from http://github.com/jeffreyhanson/rgurobi). For more information on using this package, see our paper published in Methods in Ecology and Evolution (http://dx.doi.org/doi:10.1111/2041-210X.12862).

## Installation

This package depends on several packages which can be difficult to install under Linux and Mac operating systems.

Linux (Ubuntu) users can install them typing the following code into the terminal:

```
sudo apt-get update
sudo apt-get install libgdal-dev
sudo apt-get install libproj-dev
sudo apt-get build-dep r-cran-rgl
```

Mac OSX users can install them using the code below.

```
brew install Caskroom/cask/xquartz
brew install gdal
Rscript -e "setRepositories(ind = 1:2); install.packages(c('rgdal', 'rgeos'))"
```

To install the [latest official version on CRAN](https://CRAN.R-project.org/package=raptr), use the following R code:

```{r, eval = FALSE}
install.packages("raptr")
```

To install the [development version on GitHub](https://github.com/jeffreyhanson/raptr), use this R code:

```{r, eval = FALSE}
if (!require("devtools"))
  install.packages("devtools")
devtools:::install_github("jeffreyhanson/raptr")
```

Once this package has been installed, you can read through the vignette for a tutorial on how to use it.

[View it here](https://rawgit.com/jeffreyhanson/raptr/master/inst/doc/raptr.html), or by running this R code:

```{r, eval = FALSE}
# open vignette in web browser
vignette("raptr", package = "raptr")
```

## Citation

```{r}
citation("raptr")
```
