---
title: "raspr: Identify Representative and Adequate Prioritisations in R"
author: "Jeffrey O. Hanson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Overview
A central aim in conservation is to maximise the long-term persistence of biodiversity. To achieve this, reserve networks must be sited in places that safeguard biodiversity patterns and processes. Here, we present a novel formulation of the reserve-selection problem to identify prioritisations that fulfill this objective. The representative and adequate sample problem (RASP) is a general multi-objective optimisation problem that can identify prioritisations that are adequate--in terms of how much of the feature is secured--and also representative--in terms of the spatial variation at locations where the feature is protected. Here, we present an R package to develop prioritisations using this formulation of the reserve-selection problem. We examine the behaviour of this formulation using several simulated species and a case-study species.

## Problem formulation

## Classes

## Simulated examples
### Data
To investigate the behaviour of the problem, we will generate prioritisations for three simulated species. To reduce computational effort, we will use the unreliable formulation of the problem. The first species (termed 'constant') will represent a hyper-generalist. This species will inhabit all areas with equal probability. The second species (termed 'normal') will represent a species with a single range core. The third species (termed 'bimodal') will represent a species with two distinct ecotypes, each with their own range core. To reduce computational time for this example, we will use a $15 \times 15$ grid of square planning units.

```{r, eval=FALSE}
# make planning units
sim_pus <- sim.pus(225L)

# simulate species distributions
sim_spp <- lapply(
	c('constant', 'normal', 'bimodal'),
	sim.species,
  n=1,
  x=sim_pus,
  res=1
)
```

Let's see what these species' distributions look like. Each square represents a planning unit, and the color of each denotes square denotes the probability of the species in the planning unit.

```{r, eval=FALSE}
# change the plot parameters, so we can plot the distributions side by side
par(mfrow=c(1,3))

# constant species
plot(sim_spp[[1]], main='constant species')
lines(sim_pus)

# normal species
plot(sim_spp[[2]], main='normal species')
lines(sim_pus)

# bimodal species
plot(sim_spp[[3]], main='bimodal species')
lines(sim_pus)

# reset plot parameters
par(mfrow=c(1,1))
```

Next, we will generate a set of demand points. To understand the effects of probabilities and weights on the demand points, we will generate the demand points in geographic space. These demand points will be the centroids of the planning units. Additionally, we will use the same set of demand points for each species and only vary the weights of the demand points between species.

```{r, eval=FALSE}
# generate coordinates for pus/demand points
pu_coords <- rgeos::gCentroid(sim_pus, byid=TRUE)

# calculate weights
sim_dps <- lapply(
	sim_spp,
	function(x) {
		return(extract(x, pu_coords))
	}
)

# create demand point objects
sim_dps <- lapply(
	sim_dps,
	function(x) {
		return(
			DemandPoints(
				SimplePoints(pu_coords@coords),
				c(x)
			)
		)
	}
)
```

Now, we will construct a `RaspUnsolved` object which will store all our input data and parameters to generate prioritisations.

```{r, eval=FALSE}
## create RaspOpts object
# this stores parameters for the problem (eg. BLM)
sim_ro <- RaspUnreliableOpts()

## create GurobiOpts object
# this stores parameters for solving the problem (eg. MIPGap)
sim_go <- GurobiOpts()

## create RaspData object
# create data.frame with species targets
# area targets at 20% and space targets at 0%
spp_targets <- data.frame(
	area.target=rep(0.2, 3),
	space.target=rep(0, 3),
	name=c('constant', 'normal', 'bimodal')
)

# calculate probability of each species in each pu
pu_probabilities <- calcSpeciesAverageInPus(sim_pus, stack(sim_spp))

## create AttributeSpace object
# this stores the coordinates of the planning units in an attribute space
# and the coordinates and weights of demand points in the space
attr_space <- AttributeSpace(
	SimplePoints(pu_coords@coords),
	sim_dps
)

# generate boundary data information
boundary <- calcBoundaryData(sim_pus)

## create RaspData object
# this store all the input data for the prioritisation
sim_rd <- RaspData(
	sim_pus@data,
	spp_targets,
	pu_probabilities,
	list(attr_space),
  boundary,
  SpatialPolygons2PolySet(sim_pus)
)

## create RaspUnsolved object
# this store all the input data and parameters needed to generate prioritisations
sim_ru_area <- RaspUnsolved(sim_ro, sim_go, sim_rd)
```

### Single-species prioritisations
Let's keep things simple to start off with. First, we will generate a set of prioritisations that secure 20% of the species range for minimal cost.

```{r, eval=FALSE}
# make prioritisations for each species using area targets
sp1_area <- solve(spp.subset(sim_ru_area, 'constant'))
sp2_area <- solve(spp.subset(sim_ru_area, 'normal'))
sp3_area <- solve(spp.subset(sim_ru_area, 'bimodal'))

# plot prioritisations
plot(sp1_area)
plot(sp2_area)
plot(sp3_area)
```

Here, we can see that the optimal prioritisation for the uniform species is just a random selection of XXX planning units. This is because all planning units have equal probability of occupancy and so XXXX can be used to secure 20% of its distribution. The prioritisation for the normal species focuses on the planning units in the species' range core. This prioritisation, however, fails  to secure any peripheral parts of the species' distribution. Finally, the prioritisation for the bimodally distributed species only selects planning units in the bottom left corner of the study area. As a consequence, this prioritisation only preserves individuals belonging to one eco-type. All of these priotitisations fail to secure a representative portion of the species' geographic distributions because they only prioritise the parts of the species' ranges with the highest occupancy probabilities.

Now, we will generate prioritisations for each species that aim to preserve a rerepresent different parts of the species' geographic distributions explicitly consider their

```{r, eval=FALSE}
# set 20% space targets in sim_ru
sim_ru_space<-update(sim_ru_area, space.targets=0.2)

# make prioritisations for each species
sp1_space <- solve(spp.subset(sim_ru_space, 'constant'))
sp2_space <- solve(spp.subset(sim_ru_space, 'normal'))
sp3_space <- solve(spp.subset(sim_ru_space, 'bimodal'))

# plot prioritisations
plot(sp1_space)
plot(sp2_space)
plot(sp3_space)
```

Here, we can see that the prioritisations that explictly consider the space targets do a much better job at preserving different parts of the species' distributions. Before, when only considering area targets, the prioritisation for the constant species was seemingly random. But now, the prioritisation is spread out across the species' distribution. The prioritisaiton for the normal species now captures the core distribution of the species and also parts of its range margin. Lastly, the prioritisation for the bimodally species captures both of eco-types. These results show that the inclusion of space targets in the reserve selection process can significantly change the resulting prioritisation.

### Multi-species prioritisations
Previously, we have generated prioritisations for only one species at a time. However, real-world prioritisations are often generated for multiple-species. Here, we will generate prioritisations for all three of the simulated species. First, we will generate a prioritisation that only aims to preserve 20% of the area they occupy. Then, we will generate a prioritisation that also aims to preserve a representative portion of their geographic distribution. After doing this we will compare the two prioritisations.

```{r, eval=FALSE}
# make prioritisations
sim_mrs_area<-solve(sim_ru_area)
sim_mrs_space<-solve(sim_ru_space)

# plot prioritisations
plot(sim_mru_area)
plot(sim_mru_space)
```

Here we can see that the inclusion of space-based targets to ensure that prioritisations secure a representative portion of the species' range changes the both the selection of planning units and also the number of planning units that are selected. The area-based prioritistion is comprised of `r sum(sim_mrs_area@results@selections[1,])` units, and the space-based prioritisation is comprised of `r sum(sim_mrs_space@results@selections[1,])` units. These simulated species demonstrate that under the simplest of distributions the inclusion of space-based targets can result in different prioritisations. However, real world planning exercises may use attribute spaces that have a complex relationship with geographic space, and this will affect the difference of including space-based targets.

## Case-study examples
### Overview
Here we will investigate how including space-based targets can affect prioritisations using a real-world dataset. We will generate prioritisations for the pale-headed rosella (_Platycercus adscitus_). This Australian bird species persists across broad environmental gradients across north-eastern Australia. Firstly, we will generate a typical area-based prioritisation that aims to capture 20% of its distribution. Then we will generate a prioritisation that aims to capture the neutral evolutionary processes acting on this species. To do this we will generate a prioritisation that secures indiviudals in different parts of its geographic range. Next, we will generate a prioritisation that aims to capture the adaptive evolutionary processes acting on this species, by securing individuals in different environmental conditions. Lastly, we will also generate a prioritisation that secures both of these processes and see how this prioritisations compares to the previous prioritisations.

### Data
Survey data for this species was obtained from [BirdLife Australia](http://birdlife.org.au/projects/atlas-and-birdata). This was used to generate a probability map (1km^2 resolution) of its distribution across Australia. To improve the accuracy of this map, areas well outside of its known distribution were set to 0. This was acheived by masking out [biogeographic regions](https://www.environment.gov.au/land/nrs/science/ibra) where the species was not observed and did not have a neighboring region where the species was observed. The resulting map is stored in the `cs_spp` object.

```{r, eval=FALSE}
# load data
data(cs_spp)

# plot species distribution
plot(cs_spp)
```

Planning units (10km^2 resolution) were generated across Australia, and then clipped to the Queensland state borders and coastline. To ensure our prioritisations include information on the distribution of existing protected areas, [Australia's current protected area network](http://www.protectedplanet.net/) were intersected with the planning units. Planning units with more than 50% of their area inside a protected area had their status set to two (following conventions in Marxan). Since we do not have cost data, the prioritisations will aim to select the minimum number of planning units required to meet the targets. The planning units are in the `cs_pu` object.

```{r, eval=FALSE}
# load data
data(cs_pu)

## plot planning units
# denote units not inside a protected area with white
plot(cs_pu[which(cs_pu$status==0),], col='white')

# denote units inside a protected area with green
plot(cs_pu[which(cs_pu$status==2),], col='white')
```

To map the distribution of environmental conditions across the species' range, 21 [bioclimatic layers](http://www.worldclim.org/bioclim) were obtained. These layers were cropped to Australia and subject to [detrended correspondance analysis(http://cc.oulu.fi/~jarioksa/softhelp/vegan/html/decorana.html) to produce two new variables (explaning XX percent of the variation). These layers are stored in the `cs_space` object.

```{r, eval=FALSE}
# set up plotting window
par(mfrow=c(1,2))
# plot first variable
plot(cs_space[[1]])

# plot second variable
plot(cs_space[[2]])

# reset window
par(mfrow=c(1,1))
```

### Prioritisations
To simplify the process of formatting data and generating prioritisations, we can use the `rasp` function. First, we will generate an area-based prioritisation that aims to capture 20% of the rosella's range. We will use 100 demand points to map the geographic and environmental spaces.

```{r, eval=FALSE}
# make area-based prioritisation
cs_rs_area <- rasp(cs_pus, cs_species, cs_spaces, area.targets=0.2, space.targets=0.2, n.demand.points=100L, include.geographic.space=TRUE, formulation='unreliable')

# plot prioritisation
plot(cs_rs_area)

# plot prioritisation in environmental space
space.plot(cs_rs_area)
```

Next, we will update the space targets


### Fragmentation
